NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2019-04-18 Manuscript/manuscriptPGPE.tex
Update text

\section{Molecular dynamics simulations}

Molecular dynamics simulation data were collected using
the Open Collaboration method \cite{botan15}, with
the NMR\-lipids Project blog (\url{nmrlipids.blogspot.fi}) and
GitHub repository (\url{github.com/NMRlipids/NMRlipidsIVotherHGs})
as the communication platforms.
The simulated systems are listed in 
Table \ref{systems} (pure PE and PG bilayers without additional ions).
%and \ref{mixedIONsystems} (mixed PC:PS bilayers at various salt concentrations).
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

@@

@@

\section{Results and Discussion}

@@

The glycerol backbone and $\alpha$-carbon peaks in INEPT spectra of POPE were assigned based on
previously measured POPC spectra (Fig. \ref{POPEspectra}) \cite{ferreira13}. The $\beta$-carbon peak was assigned based
on $^{13}$C chemical shift table for amines available at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
The order parameters for the glycerol backbone and headgroup C--H bonds were determined
from 2D-RPDLF and S-DROSS experiments (Fig. \ref{POPEspectra}), as described previously \cite{NMRlipidsIVps}.
The POPE experiments were recorded at 310~K, where the bilayer is in liquid disordered phase \cite{??}.
\todo{To be checked by Tiago}.

@@

\subsection{PE headgroup and glycerol backbone}
%\todo{Add results from this discussion: \url{http://nmrlipids.blogspot.com/2017/03/nmrlipids-iv-headgroup-glycerol.html?showComment=1502290977005#c4964787551006026647}}
Order parameters from Slipids simulations 
and experiments for DPPE are shown in Fig. \ref{HGorderParametersPE}.
Glycerol backbone order parameters in Slipids are off from experiments,
as already observed previously for PC lipids \cite{botan15}.
Order parameter signs for PE are not experimentally measured yet.
For headgroup the signs are set to give best agreement with simulations
and for glycerol to be consistent with experimental signs for PC.
Order parameter for $\beta$ carbon shows apparent agreement
with experiments. However, the sign of beta order parameter is positive,
in contrast to PC where negative sign was measured. Thus, the the beta order
parameter agrees with experiment with the assumption that its sign is opposite
than for PC. This is yet to be confirmed by experiments.
Order parameter for $\alpha$ carbon is too close to zero, even if the sign
would be correct.

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPE.pdf}
  \caption{\label{HGorderParametersPE}
    Order parameters for DPPE headgroup and glycerol
    backbone from simulations with Slipids \cite{??} and experiments
    (DPPE from \cite{seelig76} and E.coliPE from \cite{gally81}).
    Absolute values are shown, because signs are not known experimentally.
  }
  \todo{Results from united atom simulations yet to be added} \\
\end{figure}

@@

\subsection{PG headgroup and glycerol backbone}
Comparison between experiments and simulations for PG lipids is shown
in Fig. \ref{HGorderParametersPOPG}. The signs are not yet measured
experimentally. They are set to give the best argeement with experiments.
This would suggest that the $\beta$ order parameter would be positive,
in contrast to PC and PS headgroups, were negative signs were measured.
Even thought the signs turned out to be correct, the tested models would
not give a very good argeement with the experiments.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPG.pdf}
  \caption{\label{HGorderParametersPOPG}
    Order parameters for PG headgroup and glycerol backbone from simulations and experiments without CaCl$_2$ 
    (POPG from \cite{borle85} contains 10mM of PIPES, DPPG from \cite{wohlgemuth80} contains 10mM PIPES and 100mM CaCl,
    E.Coli PG results from \cite{gally81}).
    Signs are not known for experimental data. They are determined to give best agreement
    with simulations. This is not reliable and should be corrected when experimental
    data becomes available.
  }
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CIdensPG.eps}
  \caption{\label{CIdensPG}
    Counterion densities from simulations with PG lipids.
  }
\end{figure}

\subsection{Lipid headgroup interactions in PC:PE and PC:PG mixtures}
The headgroup 
order parameters increase with the addition of negatively charged PS and PG lipids,
decrease when mixed with positively charged surfactants and
are less affected by the addition of zwitterionic PE lipids or cholesterol.
In addition to the results summarized in Fig. \ref{HGorderparametersPCvsPEPSPGchol},
also mixtures of PC with negatively charged PI, CL, PA, and zwitterionic SM
follow the eletrometer concept \cite{scherer87}.
to mixtures of differently charged
lipids are collected from different experiments in Fig. \ref{HGorderparametersPCvsPEPSPGchol}.
%\section{PC lipid headgroup response to different mixtures in experiments}
As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
surfactants. The changes with the addition of neutral lipids is significantly smaller.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
%  \caption{\label{HGorderparametersPCvsPEPSPGchol}
%    PC headgroup order parameters from experiments of mixtures with
%    PE, PS, PG and cholesterol \cite{scherer87,scherer89,ferreira13}.
%    Signs are determined as discussed in \cite{botan15,ollila16}.
%  }
%\end{figure}

The headgroup order parameters for PC lipids (POPC and DOPC)
mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
from different simulation model and experiments \cite{scherer87} with different
mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is not the case in simulation data shown in Fig. \ref{HGorderparametersPCvsPEPSPG}.
The addition of DOPE into a POPC and DOPC bilayers significantly decreases the PC headgroup
order parameters in simulations with OPLS compatible version of the Berger force field \cite{tieleman06}
in contrast to experiments \cite{scherer87}. On the other hand, the increase of the PC
headgroup order parameters in CHARMM36 simulations mixed with PS and PG lipids is significantly
smaller than in experiments.

\todo{This is text by P. Fuchs, copied from the blog.}
Area results in nm$^2$, the error is <= 0.003 nm$^2$
- pure POPC
CHARMM36: 0.624
Berger : 0.649
- POPC/POPE 50:50
CHARMM36 : POPC 0.609, POPE 0.557
Berger-hacked: POPC 0.637, POPE 0.632
-----
One can see that CHARMM 36 predicts a drop in the area on going from pure POPC to POPC/POPE 50:50. This means that POPC pack tightly to POPE.
In contrast, the values for Berger are not that changed.
The POPE value predicted by CHARMM 36 (in the mixture POPC/POPE 50:50) is much smaller than that predicted by Berger.
\todo{The experimental acyl chain order parameters for POPE \cite{pare98} seem larger than reported for POPC \cite{ferreira13},
  which supports the more condensed PE bilayer. In principle, this is beyond the scope of this work (lipid headgroups), but we can
consider mentioning this.}

\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPSPG}
    PC headgroup order parameters from mixtures with PE, PS and PG
    lipids with various mole fractions from different simulation models and experiments \cite{scherer87}.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure*}

%\subsection{Effect of PC on PS and PG headgroups}

The headgroup order parameters for PS and PG lipid mixtures with PC
having different mole fractions from simulations and experiments \cite{borle85,roux90}
are shown in Fig. \ref{HGorderparametersPSPGvsPC}. The effect of increasing
amount of PC to PS headgroup seems to qualitatively incorrect in CHARMM36 simulations.
The $\beta$-carbon order parameter increases in experiment, but decreases
in simulations with both tested counterions (Na+ and K+). Larger
$\alpha$-carbon order parameter decreases with the addition of PC
in experiment, while the lower remains unchanged. In simulations the
larger increases and the lower decreases.
Interestingly, the $\alpha$-carbon order parameters are closer to experiments
in pure PS system with K+ counterions than with Na+. 
The changes in PG headgroup order parameters are minor in simulations, which
is in line with the only available experiment for the $\beta$-carbon.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPGvsPC.eps}
  \caption{\label{HGorderparametersPSPGvsPC}
    PS and PG lipid headgroup order parameters from mixtures with PC
    lipids with various mole fractions from different simulation models and experiments \cite{borle85,roux90}.
    Signs are not yet known experimentally for PG, thus the signs give by simulations are used.
    Signs for PS are measures as described in SI.
  }
  \todo{Some simulations contain potassium as counterions, while some sodium.
    All experiments here contain some amount of sodium salt. The best
    ion concentrations for comparison should be figured out.} \\
  \todo{Why there is difference between CHARMM36 simulation results from POPS:POPC mixture and
  pure POPS? Discussion in https://github.com/NMRLipids/NMRlipidsIVotherHGs/issues/1}
\end{figure}

\subsection{Ion binding to PE lipid bilayers}

\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/NAdensPE.eps}
  \caption{\label{NAdensPE}
    Sodium (solid line) and choride ion density profiles along membrane normal
    from different simulations with PE lipids.
  }
\end{figure}

\subsection{Ca$^{2+}$ binding in bilayers with negatively charged PG and PS lipids}

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is linearly proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
This molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Figs \ref{OrderParametersWithCaCl},
\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
%  \caption{\label{OrderParametersWithCaCl}
%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
%  }
%  \todo{Check the NaCl concentrations in the samples.}
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
%  \caption{\label{OrderParametersWithCaClBELOW1M}
%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
%  }
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
%  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
%    The change of PC headgroup order parameters
%    in the presence of different amount of negatively charged lipids
%    respect to the values without added CaCl$_2$.
%    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
%  }
%\end{figure}

PC headgroup order parameters increase when negatively charged
PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
as expected based on electrometer concept \cite{seelig87}
(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.
Further, the order parameters decrease with the addition
of CaCl$_2$ and the decrease is more pronounced for systems with more
negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
the Ca2+ binding presumably fully cancels the charge from negative lipids and
overcharging occurs above these concenterations.
The interpretation of this data and some other results has been that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
    (ii) the free Ca$^{2+}$
    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
    charge, either due to protein or to lipid; \\
    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
    of bound Ca$^{2+}$ at neutral and charged lipids; \\
    (iv) the actual binding step can be described by a Langmuir
    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
    used to describe the chemical binding equilibrium.''}
\end{displayquote}

Comparison of Ca$2+$ binding in PG between CHARMM36 simulations and experiments \cite{borle85}
is shown in Fig. \ref{changesWITHCaClPG}. The decrease of $\alpha$ order parameter
is in agreement with experiments, while decerase of $\beta$ order parameter is overestimated.
The result is very similar to the results with PC in NMRlipids II publication \cite{catte16}.
It should be, however, noted that the $\beta$-order parameters are not actually measured for PG,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
parameter of bound charge
\todo{The response of CHARMM36 to cationic surfactant against experiments \cite{scherer89} to be checked.
  I have already ran the simulations, analysis to be done.}.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG.eps}
  \caption{\label{changesWITHCaClPG}
    PG order parameters as a function CaCl$_2$ concentration from experiments \cite{borle85} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured. 
  }
\end{figure*}

Also dependence of $\beta$-carbon of PG on CaCl$_2$ concentration is compared with
experiments \cite{borle85} in Fig. \ref{PSPGchangesWITHCaCl}. Absolute value of
the order parameter is too large without ions, but rapid decrease due to addition of
CaCl$_2$ is observed in agreement with experiments for systems with 1:1 mixture of POPC and POPG.
In addition, absolute value in systems with CaCl$_2$ is in agreement with experiments.
However, system with 4:1 mixture of POPC and POPG behaves differently, but experimental
data is not available for comparison for this mixture.

\todo{More simulation data for systems with negatively charged lipids and CaCl$_2$ to be collected}

\section{Effect of Ca$2+$ binding on PG and PE headgroup}

Also the experimental order parameters for PS and PG headgroups
as a function of CaCl$_2$ concentration are shown in Fig. \ref{PSPGchangesWITHCaCl}.
%\todo{These should be compared to simulations for potential structural interpretation of the changes.}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/PGwithCaCl.eps}
%  \caption{\label{PSPGchangesWITHCaCl}
%    PG and PS order parameters a function CaCl$_2$ concentration taken from \cite{borle85} and \cite{roux90}, respectively.
%  }
%  \todo{Get the small concentration data from the inserts}
%\end{figure}

@@

>>>>

\subsection{Molecular dynamics simulations}

Molecular dynamics simulation data were collected using
the Open Collaboration method \cite{botan15}, with
the NMR\-lipids Project blog (\url{nmrlipids.blogspot.fi}) and
GitHub repository (\url{github.com/NMRlipids/NMRlipidsIVotherHGs})
as the communication platforms.
The simulated systems are listed in 
Tables \ref{systems} (pure PE and PG bilayers without additional ions).
and \ref{systemsII} (mixtures and systems with additional ions).
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

@@

  }\label{systemsII}

@@

\clearpage
\section{Results and Discussion}

@@

The glycerol backbone and $\alpha$-carbon peaks in INEPT spectra of POPE were assigned based on
previously measured POPC spectra (Fig. \ref{POPEspectra}) \cite{ferreira13}. The $\beta$-carbon peak was assigned based
on $^{13}$C chemical shift table for amines available at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
The order parameters for the glycerol backbone and headgroup C--H bonds were determined
from 2D-RPDLF and S-DROSS experiments (Fig. \ref{POPEspectra}), as described previously \cite{NMRlipidsIVps}.
The POPE experiments were recorded at 310~K, where the bilayer is in liquid disordered phase \cite{??}.
\todo{Details to be checked by Tiago}.

@@

\subsection{Headgroup and glycerol backbone of POPE and POPG in MD simulations}

The glycerol backbone order parameters in Slipid simulations are off from experiments
for both PE and PG lipids (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
as already reported previously for PC lipids \cite{botan15} and PS lipids \cite{NMRlipidsIVps}.
The Slipid simulations were able to capture the essential differences between PC and PS lipid
headgroups \cite{NMRlipidsIVps}, but this is not the case for PE and PG lipids.
 \begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPE.pdf}
  \caption{\label{HGorderParametersPE}
    The headgroup and glycerol backbone order parameters of PE lipids
    from experiments (POPE and signs this work, DPPE from Ref.~\citenum{seelig76}
    and E.coliPE from Ref.~\citenum{gally81}) and simulations with different force fields.
  }
  \todo{Results from united atom GROMOS,CHARMM36ua,OPLS-UA,Berger and GROMOS-CKP simulations yet to be analyzed and added} \\
\end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPG.pdf}
  \caption{\label{HGorderParametersPOPG}
    The headgroup and glycerol backbone order parameters of PG lipids
    from experiments (POPG and signs from this work and from Ref.~\citenum{borle85}, %contains 10mM of PIPES,
    DPPG with 100mM NaCl from Ref.~\citenum{wohlgemuth80},% contains 10mM PIPES and,
    and E.Coli PG results from Ref.~\citenum{gally81}).
    and simulations with different force fields.
  }
\end{figure}

@@

CHARMM36 simulations seems to capture all the essential differences between
headgroup order parameters of PC PE, PG (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
as previously observed also for PS lipids \cite{NMRlipidsIVps}. Therefore,
the CHARMM36 simulations can be used to analyze the structural differences between
headgroups \todo{We should do such analysis. Maybe similar to the PS part \cite{NMRlipidsIVps}?}

\subsection{Lipid headgroup interactions in PC:PE and PC:PG mixtures}
According to the electrometer concept, the headgroup order parameters increase
with the addition of negatively charged PG lipids, but
%decrease when mixed with positively charged surfactants and
are not affected by the addition of zwitterionic PE lipids (Fig.~\ref{HGorderparametersPCvsPEPG}) \cite{seelig87, scherer87}. % or cholesterol.
%In addition to the results summarized in Fig. \ref{HGorderparametersPCvsPEPSPGchol},
%also mixtures of PC with negatively charged PI, CL, PA, and zwitterionic SM
%follow the eletrometer concept \cite{scherer87}.
%to mixtures of differently charged
%lipids are collected from different experiments in Fig. \ref{HGorderparametersPCvsPEPSPGchol}.
%%\section{PC lipid headgroup response to different mixtures in experiments}
%As shown in Fig. \ref{HGorderparametersPCvsPEPSPGchol}, order parameters of PC
%headgroup behave in various lipid mixtures as expected from the electrometer concept \cite{seelig87, scherer87},
%i.e., order parameters increase when anionic lipids are mixed with PC and decrease with cationic
%surfactants. The changes with the addition of neutral lipids is significantly smaller.
\begin{figure*}[]
  \centering
  \includegraphics[width=16.0cm]{../Figs/HGorderparametersPCvsPEPG.eps}
  \caption{\label{HGorderparametersPCvsPEPG}
    Modulation of PC headgroup order parameters with increasing amount of PE (left) and PG (right) in bilayer
    from experiments \cite{scherer87,macdonald87} and simulations with different force fields.
    Signs are determined as discussed in \cite{botan15,ollila16}.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.}
\end{figure*}
%The headgroup order parameters for PC lipids (POPC and DOPC)
%mixed with PE, PS and PG lipids are shown in Fig. \ref{HGorderparametersPCvsPEPSPG}
%from different simulation model and experiments \cite{scherer87} with different
%mole fractions. As already discussed previosly, the PC lipid headgroup behaviour
%follows the electrometer concept in experiments when mixed with other lipids, i.e., the order
%parameters increase when mixed with negatively charged lipids (PS, PI, CL, PA and PG)
%remains almost unchaged when mixed with neutral lipids (PE and SM) \cite{scherer87}.
This is roughly observed also in simulations here, althouht the changes in PC headgroup with
PE are slightly overestimated and changes with PG lipid are slightly underestimated.
The underestimated increase of PC headgroup order parameters with the addition of
negaticely charged lipids was previously interepreted to indicate overestimated
counterion binding affinity, which overcompensates the effect of negative charge \cite{NMRlipidsIVps}.
Thus, the counterion binding to PG lipid bilayers (Fig. \ref{CIdensPG}) may be slightly
overestimated also in simulations here.
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CIdensPG.eps}
  \caption{\label{CIdensPG}
    Counterion densities from simulations with PG lipids.
  }
\end{figure}

\todo{This is text by P. Fuchs, copied from the blog. \\
Area results in nm$^2$, the error is $<=$ 0.003 nm$^2$ \\
- pure POPC \\
CHARMM36: 0.624 \\
Berger : 0.649 \\
- POPC/POPE 50:50 \\
CHARMM36 : POPC 0.609, POPE 0.557 \\
Berger-hacked: POPC 0.637, POPE 0.632 \\
----- \\
One can see that CHARMM 36 predicts a drop in the area on going from pure POPC to POPC/POPE 50:50. This means that POPC pack tightly to POPE. 
In contrast, the values for Berger are not that changed. 
The POPE value predicted by CHARMM 36 (in the mixture POPC/POPE 50:50) is much smaller than that predicted by Berger.\\
-------\\
The experimental acyl chain order parameters for POPE \cite{pare98} seem larger than reported for POPC \cite{ferreira13},
  which supports the more condensed PE bilayer. In principle, this is beyond the scope of this work (lipid headgroups), but we can
consider mentioning this.}

The order parameter of headgroup $\beta$-carbon of PG slightly increases with the addition
of PC \cite{macdonald87}, althought also smaller changes have been observed~\cite{borle85}
(Fig. \ref{HGorderparametersPSPGvsPC}).
The increase in PG headgroup order parameter is not observed in simulations, which
may arise from overestimated counterion binding overcompensating the changes in
electrostatic environment as suggested for PC headgroup above. However, the changes
are quite small and not fully consistent in experiments.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPGvsPC.eps}
  \caption{\label{HGorderparametersPGvsPC}
    Modulation of PG lipid headgroup order parameters with the increasing amount of PC in lipid bilayer
    from experiments \cite{borle85,macdonald87} and simulations with different force fields.
  }
\end{figure}

\clearpage
\subsection{Cation binding to PE and PG lipid bilayers}
The headgroup order parameters of PC lipids can used to measure ion binding
affinity to lipid bilayers, because their magnitude is linearly proportional
to the amount of bound charge in bilayer according to the molecular
electrometer concept \cite{seelig87,catte16}. The molecular electrometer concept
can be used also for bilayers containing PC lipids mixed with charged
lipids \cite{borle85,macdonald87,roux90,NMRlipidsIVps}.
%This is demonstrated in Figs \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and \ref{OrderParameterCHANGESWithCaClBELOW1M},
%showing order parameters for PC headgroup $\alpha$ and $\beta$ carbons
%as a function of CaCl$_2$ concentration in the presence of different amounts of
%negatively charged PS or PG lipids.
%
%PC headgroup order parameters increase when negatively charged
%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
%as expected based on electrometer concept \cite{seelig87}
%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%Further, the order parameters decrease with the addition
%of CaCl$_2$ and the decrease is more pronounced for systems with more
%negatively charged lipids (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
%At CaCl$_2$ concentrations ($\sim$ 50-300mM) where order parameters reach the values for pure PC,
%the Ca2+ binding presumably fully cancels the charge from negative lipids and
%overcharging occurs above these concenterations.
Based on the electrometer concept and other data is has been suggested that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
    (ii) the free Ca$^{2+}$
    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
    charge, either due to protein or to lipid; \\
    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
    of bound Ca$^{2+}$ at neutral and charged lipids; \\
    (iv) the actual binding step can be described by a Langmuir
    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
    used to describe the chemical binding equilibrium.''}
\end{displayquote}
The electrometer concept has been very useful in evaluating ion binding affinity in simulations
against experiments, because the headgroup order parameter changes as a function of ion concentration
can be directly compared with experiments \cite{catte16,melcr18,NMRlipidsIVps}.

Sodium binding to PC lipid bilayers is significantly overestimated by most simulations models \cite{catte16}.
Also, sodium binding to PS lipids seems to be overestimated, although the
presence of counterions complicate the comparison for negatively charged lipids \cite{NMRlipidsIVps}.
Sodium binding to PE lipids is weaker than to PC lipids in previous work \cite{catte16} in both CHARMM36 and Slipids simulations (Fig. \ref{NAdensPE}).
This cannot be evaluated using electrometer concept because experimental data with PE lipids is not
available \todo{Other experimental data should be checked and discussed}.
The difference in Slipids may also be due to different ion model used \todo{The used ion model here should be checked}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/NAdensPE.eps}
  \caption{\label{NAdensPE}
    Sodium (solid line) and choride ion density profiles along membrane normal
    from different simulations with PE lipids.
  }
\end{figure}

Calcium binding affinity to PC and PS lipid bilayers was not correctly described by
any of the standard MD simulation forced fields \cite{catte16,NMRlipidsIVps}, while
recently introduced force field with electronic continuum correction (ECC) performed better \cite{melcr18}.
The decrease of $\alpha$-carbon order parameter of PC lipids in PC:PG mixtures as a function
of calcium concentration is close to experiments CHARMM36 simulations (Fig. \ref{changesWITHCaClPG}),
but the decrease of $\beta$-carbon order parameter seems to be overestimated.
However, the $\beta$-carbon order parameter was not actually measured from these samples,
but they are calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}.
The result is similar to the $\sim$200~ns simulations with PC lipids in previous work \cite{catte16}.
However, when simulation was continued for $\mu$s, the binding affinity substantially increased
and interpretation was that calcium overbinds to PC lipid in CHARMM36. Therefore, the
conclusion seems to be similar here, although the new NBfix parameters may complicate
the situation \todo{The status of NBfix parameters in these simulations should be checked.}.
%Anyway, the data presented in NMRlipids II project and in Fig. \ref{changesWITHCaClPG} together
%suggest that Calcium binding is similarly overestimated by CHARMM36 model in pure POPC bilayers and mixtures with
%POPG. The good agreement of $\alpha$ carbon would be explained by too weak dependence of its order
%parameter of bound charge
%\todo{The response of CHARMM36 to cationic surfactant against experiments \cite{scherer89} to be checked.
%  I have already ran the simulations, analysis to be done.}.
\begin{figure*}[]
  \centering
  \includegraphics[width=18.0cm]{../Figs/CHANGESwithCaClPG.eps}
  \caption{\label{changesWITHCaClPG}
    (left) The headgroup order parameters of PC from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85,macdonald87} and CHARMM36 simulations.
    Note that beta order parameter is calculated from empirical relation $\Delta S_{\beta}=0.43\Delta S_{\alpha}$ \cite{akutsu81}, not actually measured.
    (right) The headgroup order parameters of PG from PC:PG mixtures as a function CaCl$_2$
    concentration from experiments \cite{borle85} and CHARMM36 simulations.
  }
\end{figure*}

The $\beta$-carbon order parameter of PG exhibits a rapid decrease with small CaCl$_2$ concentrations
and a more modest decrease with larger concentrations in experiments \cite{borle85} (Fig. \ref{PSPGchangesWITHCaCl}).
The rapid decrease with CaCl$_2$ is observed but overestimated in CHARMM36 simulation
with POPC:POPG 1:1 mixture, but not in 4:1 mixture \todo{This is little bit weird, should be checked.}.

@@

\todo{We need PC:PG simulations with CaCl$_2$ from different force fields to finish the discussion.}
